{ \sffamily \LARGE Electronic supplementary material}\\{\bf \sffamily \LARGE Evolution of reciprocity with limited payoff memory

### 2024-03-18 paper/main.tex
CH: Revising the main text

\begin{abstract}
\noindent
Direct reciprocity is a mechanism for the evolution of cooperation in repeated social interactions.
According to this literature, individuals naturally learn to adopt conditionally cooperative strategies if they have multiple encounters with their partner.
Corresponding models have greatly facilitated our understanding of cooperation, yet they often make strong assumptions on how individuals remember and process payoff information.
For example, when strategies are updated through social learning, it is commonly assumed that individuals compare their average payoffs.
This would require them to compute (or remember) their payoffs against everyone else in the population.
To understand how more realistic constraints influence direct reciprocity, we consider the evolution of conditional behaviors when individuals learn based on more recent experiences.
Even in the most extreme case that they only take into account their very last interaction, we find that cooperation can still evolve.
However, such individuals adopt less generous strategies, and they tend to cooperate less often than in the classical setup with average payoffs.
Interestingly, once individuals remember the payoffs of two or three recent interactions, cooperation rates quickly approach the classical limit.
These findings contribute to a literature that explores which kind of cognitive capabilities are required for reciprocal cooperation.
While our results suggest that some rudimentary form of payoff memory is necessary, it already suffices to remember a few interactions.
\end{abstract}

~\\
{\it Keywords:} Evolution of cooperation; direct reciprocity; repeated prisoner's dilemma; social learning; evolutionary dynamics

@@

Evolutionary game theory describes the dynamics of populations when an individual's fitness depends on the traits or strategies of other population members~\cite{hofbauer1998evolutionary, nowak:Nature:2004, hauert2005game,Traulsen:PhilTrans:2022}.  
This theory can be used to describe the dynamics of animal conflict~\citep{maynard-smith:Nature:1973}, cancer cells~\citep{Stein:PTRS:2023}, and of cooperation~\citep{nowak:Science:2006}. 
Respective models translate strategic interactions into games~\cite{smith1982evolution}. 
These games specify how individuals (players) interact, which strategies individuals can choose, and what fitness consequences (or payoffs) the different strategies have. 
In addition, these models also specify the mode by which successful strategies spread over time. 
In models of biological evolution, individuals with a high fitness produce more offspring; in models of cultural evolution, such individuals are imitated more often. 
Although biological and cultural evolution are sometimes treated as equivalent, there can be important differences~\citep{Wu2015,Smolla:PTRS:2021,Denton:TPB:2022}. 
For example, models of biological evolution do not require individuals to have any particular cognitive abilities.
Here, it is the evolutionary process itself that biases the population towards strategies with higher fitness. 
In contrast, in models of cultural evolution, individuals need to be aware of the different strategies present in the population, and they need to identify those strategies with a higher payoff. 
As a consequence, evolutionary outcomes may depend on how easily different behaviors can be learned~\citep{Chatterjee:JTB:2012}, and on how easy payoff comparisons are.

@@

These difficulties to learn strategies by social imitation are particularly pronounced in models of direct reciprocity. 
This literature follows Trivers' insight that individuals have more of an incentive to cooperate in social dilemmas when they interact repeatedly~\citep{trivers1971evolution}. 
In repeated interactions, individuals can condition their behavior on their past experiences with their interaction partner. 
They may use strategies such as Tit-for-Tat~\citep{rapoport:book:1965,axelrod1981evolution} or Generous Tit-for-Tat~\citep{molander:jcr:1985,Nowak1992tit} to preferentially cooperate with other cooperators. 
Such conditional strategies approximate human behavior fairly well~\citep{fischbacher:EconL:2001,Rand:TCS:2013,DalBo:AER:2019,Montero-Porras:SciRep:2022,Rossetti:ETH:2023} and they have also been documented in several other species~\citep{Carter:PRSB:2013,Schweinfurth:AnBehav:2019,Voelkl:PNAS:2015} -- although direct reciprocity is generally more difficult to demonstrate in animals~\citep{CluttonBrock:Nature:2009,Silk:CurrentBiology:2013,taborsky:CurrentBiology:2013}.
However, at the outset, it is not clear how easy it is to {\it learn} reciprocal strategies by social imitation. 
As one obstacle, even if others' strategies are perfectly observable, individuals might find it difficult to identify which ones have the highest payoff. 
After all, the payoff of a strategy of direct reciprocity is not determined by the outcome of any single round.
Rather, it is determined by how well this strategy fares over an entire sequence of rounds, against many different population members. 
In practice, such information might be both difficult to obtain and to process.

@@

Most models of direct reciprocity abstract from these difficulties~\citep{brauchli:JTB:1999,brandt:JTB:2006,ohtsuki:JTB:2007b,szolnoki:pre:2009b,imhof2010stochastic,van-segbroeck:prl:2012,grujic:jtb:2012,Martinez2012,stewart:pnas:2013,pinheiro:PLoSCB:2014,stewart:games:2015,Baek2016,hilbe:PNAS:2017,McAvoy:ProcA:2019,glynatsi:SCR:2020,Schmid:PlosCB:2022,Murase:SciRep:2022,Cooney:BMB:2022,Chen:PNASNexus:2023}. 
They just assume individuals can easily copy the strategies of others. 
Similarly, they just assume that updating decisions are based on the strategies' average (or expected) payoffs, which are based on all rounds and all interactions. 
These assumptions create a curious inconsistency in how models represent an individual's cognitive abilities. 
On the one hand, when playing the game, individuals are often assumed to have restricted memory. 
Respective studies typically assume that individuals make their decisions each round based on the outcome of the last round only~\citep[with only a few exceptions, see Refs.][]{Hauert1997,van-veelen:PNAS:2012,Stewart2016,Li:NatCS:2022,Murase:PLoSCompBio:2023a}. 
Yet when learning new strategies, individuals are assumed to remember (or compute) each others' precise average payoff across many rounds and many interaction partners. 
Herein, we wish to explore whether this latter assumption is actually necessary for the evolution of reciprocity through social imitation. 
We ask whether individuals can learn to adopt reciprocal strategies even when learning is based on payoff information from a limited number of rounds.

%% INTRO: Basic description of our model and our findings

To explore that question, we theoretically study imitation dynamics in the repeated prisoner's dilemma, using two extreme scenarios. 
The first scenario is the usual modeling approach. 
Here, individuals update their strategies based on their expected payoffs. 
We contrast this model with an alternative scenario where individuals update their strategies based on the very
last (one-shot) payoff they obtained. 
We find that individuals with limited payoff memory tend to adopt less generous strategies. 
Yet moderate levels of cooperation can still evolve. 
Moreover, as we increase the individuals' payoff memory to include the last two or three one-shot payoffs, cooperation rates quickly approach the rates observed in the classical baseline case.

%% INTRO: Summary and conclusions

% Overall, these findings suggest that while memory is important, already minimal payoff information may suffice for the evolution of direct reciprocity based on social learning. 
% They also suggest that the classical model of reciprocity (based on expected payoffs) can often be interpreted as a useful approximation to more realistic models that include cognitive constraints.

\todo{To address the reviewer's comment, we can either delete the last paragraph or rephrase the second-to-last paragraph as follows:}

To explore this question, we theoretically study imitation dynamics in the
repeated prisoner's dilemma, employing two extreme scenarios. In the first
scenario, which follows the conventional modeling approach, individuals update
their strategies based on their expected payoffs. We contrast this model with an
alternative scenario where individuals base their strategy updates solely on the
very last (one-shot) payoff they obtained. We observe that individuals with
limited payoff memory tend to adopt less generous strategies, yet moderate
levels of cooperation can still emerge. This suggests that while memory plays a
significant role, minimal payoff information may suffice for the evolution of
direct reciprocity through social learning. Moreover, increasing individuals'
payoff memory to include the last two or three one-shot payoffs leads to
cooperation rates quickly approaching those observed in the classical baseline
case. This implies that the classical reciprocity model, based on expected
payoffs, often serves as a useful approximation to more complex models
incorporating cognitive constraints.

@@

\noindent
{\bf Description of the game dynamics.} We consider a well-mixed population consisting of $N$~players.
Players are randomly matched in pairs to participate in a repeated donation game~\citep{sigmund2010calculus} with their respective co-player.
Each round, they can either cooperate (\(C\)) or defect (\(D\)). 
A cooperating player provides a benefit~\(b\) to the other player at their own cost~\(c\), with \(0 \!<\! c \!<\! b\). 
A defecting player provides no benefit and pays no cost. 
Thus, the players' payoffs in a single round are given by the matrix
\begin{align}
\bordermatrix{%
	& C & D \cr
C &\ b-c &\ -c\  \cr
D &\ b &\ 0\ \cr
} .
\end{align}
In particular, payoffs take the form of a prisoner's dilemma:
Mutual cooperation yields a better payoff than mutual defection ($b\!-\!c\!>\!0$), but each player individually prefers to defect independent of the co-player's action ($b\!>\!b\!-\!c$ and $0\!>\!-c$). 
To incorporate that individuals interact repeatedly, we assume that after each round, there is a constant continuation probability $\delta$ of interacting for another round. 
For $\delta\!=\!0$, we recover the case of a conventional (one-shot) prisoner's dilemma. 
Here, mutual defection is the only equilibrium. 
As $\delta$ increases, the game turns into a repeated game. Here, additional equilibria emerge, with some of them allowing for full cooperation~\citep{friedman:RES:1971,Akin:chapter:2016,hilbe:GEB:2015,stewart:pnas:2014}.

@@

\noindent
{\bf Description of the evolutionary dynamics.}
Herein, we assume population members update their strategies based on social learning. 
To model these strategy updates, we consider a pairwise comparison process~\citep{traulsen2007pairwise}. 
This process assumes that at regular time intervals, one population member is randomly selected, and given the chance to revise its strategy.
We refer to this player as the `learner'. 
With probability $\mu$ (reflecting a mutation rate), the learner simply adopts a random strategy (all reactive strategies have the same probability to be chosen). 
With the converse probability $1\!-\!\mu$, the learner randomly picks a `role model' from the population. 
The learner then compares its own payoff $\pi_\learner$ from the repeated game to the role model's payoff $\pi_\rolemodel$. 
The learner adopts the role model's strategy with a probability \(\varphi\) described by a Fermi function~\citep{blume:GEB:1995,szabo:PRE:1998}, 
\begin{equation} \label{Eq:rho}
    \varphi\left(\pi_\learner, \pi_\rolemodel\right) = \frac{1}{1\!+\! e^{- \!\beta\left(\pi_\rolemodel- \pi_\learner\right)}}.
\end{equation}
The selection strength parameter $\beta\!\ge\!0$ indicates how sensitive players are to payoff differences. 
For $\beta\!=\!0$, payoff differences are irrelevant, and the learner simply adopts the role model's strategy with probability one half. As the selection strength~$\beta$ increases, players are increasingly biased to imitate the role model only if it has the higher payoff. 
%In either case, we assume in the following that the role model's strategy can be identified.
%That is, if the learner decides to imitate the role model's strategy, the learner copies it exactly, as in previous works on direct reciprocity.

@@

We deviate from previous models in how we interpret the payoffs $\pi_\learner$ and $\pi_\rolemodel$, which form the basis of the pairwise comparisons in Eq.~\eqref{Eq:rho}. 
In previous work, these payoffs are taken to be the respective players' expected payoffs. 
We interpret that setup as a model with perfect payoff memory. 
There, the payoffs  $\pi_\learner$ and $\pi_\rolemodel$ represent an average over all possible repeated games the two individuals have played with all population members (\FigBaseResults, upper left panel). 
The use of expected payoffs is mathematically convenient, because explicit formulas for these payoffs are available~\citep{hofbauer1998evolutionary}.
Herein, we compare this model of perfect payoff memory to a model with limited payoff memory. 
In that model, the players' payoffs $\pi_\learner$ and $\pi_\rolemodel$ are taken to be the payoffs that each player received in their very last round prior to making social comparisons. 
That is, players only consider the very last repeated game they participated in, and there they only take into account the outcome of the very last round (\FigBaseResults, lower left panel). 
This assumption could reflect, for example, a strong recency bias in how individuals evaluate payoffs.  
In addition to this extreme case of limited payoff memory, later on we also explore cases in which players take into account the outcome of two, three, four, or more recent rounds.

@@

Both in the case of perfect and limited memory, we iterate the elementary strategy updating step described above for many time steps. 
This gives rise to a stochastic process that describes which strategies players adopt over time. 
We explore the dynamics of this process mathematically and with computer simulations.
For the results presented in the following, we assume that mutations are rare (\(\mu\!\rightarrow\! 0\)). 
This assumption is fairly common in evolutionary game theory, because it makes some computations more efficient~\citep{fudenberg:JET:2006,wu:JMB:2012,mcavoy:jet:2015}, and because the results can be interpreted more easily.  
However, in Section~3 of the \esm{} we show that our main results continue to hold for strictly positive mutation rates.

@@

\noindent
{\bf Stability of cooperative populations.}
To get some intuition for the differences between perfect and limited payoff memory, we first analyze when cooperation is stable in either scenario.
To this end, we consider a resident population in which all players but one adopt a strategy of Generous Tit-for-Tat, \gtft{} $=\!(1,1,q)$. 
The remaining mutant player adopts \alld. 
We say {\it cooperation is stochastically stable} if the single mutant is more likely to imitate the residents than vice versa. 
For simplicity, we consider a large population~($N\!\rightarrow\!\infty$) and strong selection~($\beta\!\rightarrow\!\infty$).
More general results are derived in the \esm.

@@

In the case of perfect payoff memory, it is straightforward to characterize when cooperation is stochastically stable. 
Here, we simply need to compute the players' expected payoffs. 
Because the population mostly consists of residents, and because residents mutually cooperate with each other, their expected payoff is $\pi_\gtft = b\!-\!c$. 
On the other hand, the defecting mutant only interacts with residents. 
Given the residents' strategy, the mutant receives a benefit in the first round, and in every subsequent round with probability $q$. 
As a result, the mutant's expected payoff is $\pi_\alld \!=\! (1\!-\!\delta\!+\!\delta q)b$. 
For perfect payoff memory, the requirement for cooperation to be stochastically stable reduces to the condition $\pi_\gtft > \pi_\alld$. 
This yields
\begin{equation} \label{Eq:PerfectMemory}
q < 1\!-\!\frac{c}{\delta  b}.
\end{equation}
In particular, we recover the previous observation that $q\!=\!1-c/(\delta b)$ is the maximum generosity that cooperators should have~\citep{molander:jcr:1985,Nowak1992tit,Schmid:NHB:2021}. 
Because $q\!\ge\!0$, we also conclude that cooperation can only be stable if $\delta \!>\! c/b$.
Again, this condition for the feasibility of direct reciprocity is the condition found in the literature~\citep{nowak:Science:2006}.

@@

The logic of the case with limited payoff memory is somewhat different. 
Here we need to compute how likely each player obtains one of the four possible payoffs $\{b\!-\!c, -c, b, 0\}$ in the very last round of a game, before they make social comparisons.
Because residents almost always interact with other residents, their last one-shot payoff is $\pi_\gtft = b\!-\!c$ almost surely. 
For the defecting mutant, there are two possibilities. 
({\it i})~ If the mutant's co-player happens to cooperate in the last round, the mutant receives $\pi_\alld\!=\!b$.
This case occurs with probability $1\!-\!\delta\!+\!\delta q$. 
({\it ii})~If the co-player defects in the last round, the mutant receives $\pi_\alld=\!0$. 
This occurs with the converse probability $\delta(1\!-\!q)$.
Because $b\!-\!c\!<\!b$, residents tend to imitate the mutant in the first case. 
Because $b\!-\!c\!>\!0$, mutants tend to imitate the resident in the second case. 
Cooperation is stochastically stable if the first case is less likely than the second. 
This yields the condition
\begin{equation} \label{Eq:LimitedMemory}
q < 1\!-\!\frac{1}{2 \delta}.
\end{equation}
Interestingly, this condition no longer depends on the exact payoff values $c$ and $b$. 
This independence arises because of our assumption of strong selection, in which case only the payoff ordering $b\!>\!c\!>\!0$ matters. Because $q$ is non-negative, condition~\eqref{Eq:LimitedMemory} can only be satisfied if $\delta\!>\!1/2$. That is, players need to interact in more than two rounds in expectation.

@@

By comparing the two cases, we find that payoff memory affects whether a conditionally cooperative strategy ($1,1,q$) is viable. 
With perfect memory, the maximum generosity $q$ needs to satisfy Eq.~\eqref{Eq:PerfectMemory}.
In particular, this generosity can become arbitrarily large, provided the game's benefit-to-cost ratio~$b/c$ and the continuation probability~$\delta$ are sufficiently large. 
In contrast, with limited payoff memory, the maximum generosity is bounded by one half, and it is independent of the benefit-to-cost ratio.\\

@@

\noindent
{\bf Evolutionary dynamics of reciprocity.}
To explore whether the previous static observations describe the dynamics of evolving populations, we turn to simulations.
We have run separate simulations for perfect and limited payoff memory. 
In each case, we consider both a low and a high benefit of cooperation ($b/c\!=\!3$ and $b/c\!=\!10$, respectively).  
For each simulation, we record which strategies ($y,p,q$) the players adopt over time.
\FigBaseResults{} depicts the conditional cooperation probabilities $p$ and $q$ (we omit the opening move \(y\) because we use a discount factor~\(\delta\) close to one, such that first-round behavior is largely irrelevant). 
In all simulations, we find that the players' strategies cluster in two regions of the strategy space. 
The first region corresponds to a neighborhood of \alld{} with $(p,q)\!\approx\!(0,0)$.
The second region corresponds to a thin strip of cooperative strategies with $(p,q)\!\approx\!(1,q)$. 
Within this strip, we observe that most strategies satisfy the constraints on $q$ imposed by the inequalities~\eqref{Eq:PerfectMemory} and~\eqref{Eq:LimitedMemory}. 
That is, with perfect memory, most evolving strategies have $q\!<\!1\!-\!c/b$, whereas with limited payoff memory, most strategies have $q\!<\!1/2$. 
In particular, for limited payoff memory, changes in the benefit parameter have no effect on the qualitative distribution of strategies.

@@

We complement these analytical results with additional simulations, see \FigHigherMemory.
We observe that a minimal increase in the players' payoff memory (compared to the baseline case with a single round recalled) can promote cooperation considerably. 
Specifically, in all four scenarios with extended memory, we see similar cooperation rates, and they approach the rates observed under perfect memory. 
These results suggest that while it takes {\it some} payoff memory to sustain substantial cooperation rates, the requirements on memory seem to be rather modest. 
Already remembering a few interactions, either with the same co-player or across different co-players, may provide players with enough information to adopt reciprocal strategies.

>>>>

\begin{abstract}
\noindent
Direct reciprocity is a mechanism for the evolution of cooperation in repeated social interactions.
According to this literature, individuals naturally learn to adopt conditionally cooperative strategies if they have multiple encounters with their partner.
Corresponding models have greatly facilitated our understanding of cooperation, yet they often make strong assumptions on how individuals remember and process payoff information.
For example, when strategies are updated through social learning, it is commonly assumed that individuals compare their average payoffs.
This would require them to compute (or remember) their payoffs against everyone else in the population.
To understand how more realistic constraints influence direct reciprocity, we consider the evolution of conditional behaviors when individuals learn based on more recent experiences.
Even in the most extreme case that they only take into account their very last interaction, we find that cooperation can still evolve.
However, such individuals adopt less generous strategies, and they cooperate less often than in the classical setup with average payoffs.
Interestingly, once individuals remember the payoffs of two or three recent interactions, cooperation rates quickly approach the classical limit.
These findings contribute to a literature that explores which kind of cognitive capabilities are required for reciprocal cooperation.
While our results suggest that some rudimentary form of payoff memory is necessary, it suffices to remember a few interactions.
\end{abstract}

~\\
{\it Keywords:} Evolution of cooperation; evolutionary dynamics; direct reciprocity; repeated prisoner's dilemma; social learning

@@

Evolutionary game theory describes the dynamics of populations when an individual's fitness depends on the traits or strategies of other population members~\cite{hofbauer1998evolutionary, nowak:Nature:2004, hauert2005game,Traulsen:PhilTrans:2022}.  
This theory can be used to describe the dynamics of animal conflict~\citep{maynard-smith:Nature:1973}, cancer cells~\citep{Stein:PTRS:2023}, and of cooperation~\citep{nowak:Science:2006}. 
Respective models translate strategic interactions into games~\cite{smith1982evolution}. 
These games specify how individuals (players) interact, which strategies individuals can choose, and what fitness consequences (or payoffs) the different strategies have. 
In addition, these models also specify the mode by which successful strategies spread over time. 
In models of biological evolution, individuals with a high fitness produce more offspring; in models of cultural evolution, such individuals are imitated more often. 
Although biological and cultural evolution are sometimes treated as equivalent, there can be important differences~\citep{Wu2015,Smolla:PTRS:2021,Denton:TPB:2022}. 
For example, models of biological evolution do not require individuals to have any particular cognitive abilities.
Here, it is the evolutionary process itself that biases the population towards strategies with higher fitness. 
In contrast, in models of cultural evolution, individuals need to be aware of the different strategies present in the population, and they need to identify those strategies with a higher payoff. 
As a consequence, evolutionary outcomes may depend on how easily different behaviors can be learned~\citep{Chatterjee:JTB:2012}, as well as on how readily payoffs can be compared.

@@

These difficulties to learn strategies by social imitation are particularly pronounced in models of direct reciprocity. 
This literature follows Trivers' insight that individuals have more of an incentive to cooperate in social dilemmas when they interact repeatedly~\citep{trivers1971evolution}. 
In repeated interactions, individuals can condition their behavior on their past experiences with their interaction partner. 
They may use strategies such as Tit-for-Tat~\citep{rapoport:book:1965,axelrod1981evolution} or Generous Tit-for-Tat~\citep{molander:jcr:1985,Nowak1992tit} to preferentially cooperate with other cooperators. 
Such conditional strategies approximate human behavior fairly well~\citep{fischbacher:EconL:2001,Rand:TCS:2013,DalBo:AER:2019,Montero-Porras:SciRep:2022,Rossetti:ETH:2023} and they have also been documented in several other species~\citep{Carter:PRSB:2013,Schweinfurth:AnBehav:2019,Voelkl:PNAS:2015}--although direct reciprocity is generally more difficult to demonstrate in animals~\citep{CluttonBrock:Nature:2009,Silk:CurrentBiology:2013,taborsky:CurrentBiology:2013}.
However, at the outset, it is not clear how easy it is to {\it learn} reciprocal strategies by social imitation. 
As one obstacle, even if others' strategies are perfectly observable, individuals might find it difficult to identify which ones have the highest payoff. 
After all, the payoff of a strategy of direct reciprocity is not determined by the outcome of any single round.
Rather, it is determined by how well this strategy fares over an entire sequence of rounds, against many different population members. 
In practice, such information might be difficult both to obtain and to process.

@@

Most models of direct reciprocity do not address these difficulties~\citep{brauchli:JTB:1999,brandt:JTB:2006,ohtsuki:JTB:2007b,szolnoki:pre:2009b,imhof2010stochastic,van-segbroeck:prl:2012,grujic:jtb:2012,Martinez2012,stewart:pnas:2013,pinheiro:PLoSCB:2014,stewart:games:2015,Baek2016,McAvoy:ProcA:2019,glynatsi:SCR:2020,Schmid:PlosCB:2022,Murase:SciRep:2022,Cooney:BMB:2022,Chen:PNASNexus:2023} and, instead, assume individuals can easily copy the strategies of others. 
Similarly, they assume that updating decisions are based on the strategies' average (or expected) payoffs, which are based on all rounds and all interactions. 
These assumptions create a curious inconsistency in how models represent an individual's cognitive abilities. 
On the one hand, when playing the game, individuals are often assumed to have restricted memory. 
Respective studies typically assume that individuals make their decisions each round based on the outcome of the last round only~\citep[with only a few exceptions, see Refs.][]{Hauert1997,van-veelen:PNAS:2012,Stewart2016,hilbe:PNAS:2017,Li:NatCS:2022,Murase:PLoSCompBio:2023a}. 
Yet when learning new strategies, individuals are assumed to remember (or compute) each others' precise average payoff across many rounds and many interaction partners. 
Herein, we wish to explore whether this latter assumption is actually necessary for the evolution of reciprocity through social imitation. 
We ask whether individuals can learn to adopt reciprocal strategies even when learning is based on payoff information from a limited number of rounds.

%% INTRO: Basic description of our model and our findings 
To explore that question, we theoretically study imitation dynamics in the repeated prisoner's dilemma, using two extreme scenarios. 
The first scenario is the classical modeling approach. 
Here, individuals update their strategies based on their expected payoffs. 
We contrast this model with an alternative scenario where individuals update their strategies based on the very
last (one-shot) payoff they obtained. 
We find that individuals with limited payoff memory tend to adopt less generous strategies. 
Yet moderate levels of cooperation can still evolve. 
Moreover, as we increase the individuals' payoff memory to include the last two or three one-shot payoffs, cooperation rates quickly approach the rates observed in the classical baseline.

@@

\noindent
{\bf Description of the game dynamics.} We consider a well-mixed population consisting of $N$~players.
Players are randomly matched in pairs to participate in a repeated donation game~\citep{sigmund2010calculus} with their respective co-player.
Each round, they can either cooperate (\(C\)) or defect (\(D\)). 
A cooperating player provides a benefit~\(b\) to the other player at their own cost~\(c\), with \(0 \!<\! c \!<\! b\). 
A defecting player provides no benefit and pays no cost. 
Thus, the players' payoffs in a single round are given by the matrix
\begin{equation}
\bordermatrix{%
	& C & D \cr
C &\ b-c &\ -c\  \cr
D &\ b &\ 0\ \cr
} .
\end{equation}
In particular, payoffs take the form of a prisoner's dilemma:
Mutual cooperation yields a better payoff than mutual defection ($b\!-\!c\!>\!0$), but each player individually prefers to defect, independent of the co-player's action ($b\!>\!b\!-\!c$ and $0\!>\!-c$). 
To incorporate repetition, we assume that after each round, there is a constant continuation probability $\delta$ of interacting for another round. 
For $\delta\!=\!0$, we recover the case of a conventional (one-shot) prisoner's dilemma. 
Here, mutual defection is the only equilibrium. 
As $\delta$ increases, the game turns into a repeated game. Here, additional equilibria emerge, with some of them allowing for full cooperation~\citep{friedman:RES:1971,Akin:chapter:2016,hilbe:GEB:2015,stewart:pnas:2014}.

@@

\noindent
{\bf Description of the evolutionary dynamics.}
Herein, we assume population members update their strategies based on social learning. 
To model these strategy updates, we consider a pairwise comparison process~\citep{traulsen2007pairwise}. 
This process assumes that at regular time intervals, one population member is randomly selected, and given the chance to revise its strategy.
We refer to this player as the ``learner.'' 
With probability $\mu$ (reflecting a mutation rate), the learner simply adopts a random strategy (all reactive strategies have the same probability to be chosen). 
With the converse probability $1\!-\!\mu$, the learner randomly picks a ``role model'' from the population. 
The learner then compares its own payoff $\pi_\learner$ from the repeated game to the role model's payoff $\pi_\rolemodel$. 
The learner adopts the role model's strategy with a probability \(\varphi\) described by a Fermi function~\citep{blume:GEB:1995,szabo:PRE:1998}, 
\begin{equation} \label{Eq:rho}
    \varphi\left(\pi_\learner, \pi_\rolemodel\right) = \frac{1}{1\!+\! e^{- \!\beta\left(\pi_\rolemodel- \pi_\learner\right)}}.
\end{equation}
The selection strength parameter $\beta\!\ge\!0$ indicates how sensitive players are to payoff differences. 
For $\beta\!=\!0$, payoff differences are irrelevant, and the learner simply adopts the role model's strategy with probability $1/2$. As the selection strength~$\beta$ increases, players are increasingly biased to imitate the role model only if it has the higher payoff. 
%In either case, we assume in the following that the role model's strategy can be identified.
%That is, if the learner decides to imitate the role model's strategy, the learner copies it exactly, as in previous works on direct reciprocity.

@@

We deviate from previous models in how we interpret the payoffs $\pi_\learner$ and $\pi_\rolemodel$, which form the basis of the pairwise comparisons in Eq.~\eqref{Eq:rho}. 
In previous work, these payoffs are taken to be the respective players' expected payoffs. 
We interpret that setup as a model with perfect payoff memory. 
There, the payoffs  $\pi_\learner$ and $\pi_\rolemodel$ represent an average over all possible repeated games the two individuals have played with all population members (\FigBaseResults, upper left panel). 
The use of expected payoffs is mathematically convenient, because explicit formulas for these payoffs are available~\citep{hofbauer1998evolutionary}.
Herein, we compare this model of perfect payoff memory to a model with limited payoff memory. 
In the latter model, the players' payoffs $\pi_\learner$ and $\pi_\rolemodel$ are taken to be the payoffs that each player received in their very last round prior to making social comparisons. 
That is, players only consider the very last repeated game they participated in, and there they only take into account the outcome of the very last round (\FigBaseResults, lower left panel). 
This assumption could reflect, for example, a strong recency bias in how individuals evaluate payoffs.  
In addition to this extreme case of limited payoff memory, later on we also explore cases in which players take into account the outcome of two, three, four, or more recent rounds.

@@

Both in the case of perfect and limited memory, we iterate the elementary strategy update step described above many times. 
This gives rise to a stochastic process that describes which strategies players adopt over time. 
Provided the selection strength is finite, this process satisfies the mathematical property of being ergodic. 
This implies that in the long run, the time average of the players' cooperation rates converges, and that this limit is independent of the initial population. 
We explore the dynamics of this process mathematically and with numerical simulations.
These simulations are run sufficiently long for convergence to occur (we numerically checked that independent runs are within a 1\% error tolerance). 
For the results presented in the following, we assume that mutations are rare (\(\mu\!\rightarrow\! 0\)). 
This assumption is fairly common in evolutionary game theory, both because it makes some computations more efficient~\citep{fudenberg:JET:2006,wu:JMB:2012,mcavoy:jet:2015} and because the results can be interpreted more easily.  
However, in Section~3 of the \esm{} we show that our main results continue to hold for strictly positive mutation rates.

@@

\noindent
{\bf Stability of cooperative populations.}
To get some intuition for the differences between perfect and limited payoff memory, we first analyze when full cooperation is stable in either scenario.
For a population to be fully cooperative, players need to cooperate in the first round (that is, $y\!=\!1$) and after any round with mutual cooperation (that is, $p\!=\!1$). 
We refer to such a strategy as Generous Tit-for-Tat, \gtft{} $=\!(1,1,q)$. 
To explore whether such a population consisting of \gtft{} players is stable, we introduce a single mutant who adopts \alld. 
We say {\it (full) cooperation is stochastically stable} if the single mutant is more likely to imitate the residents than vice versa. 
For simplicity, we consider a large population~($N\!\rightarrow\!\infty$) and strong selection~($\beta\!\rightarrow\!\infty$).
More general results are derived in the \esm.

@@

In the case of perfect payoff memory, it is straightforward to characterize when cooperation is stochastically stable. 
Here, we simply need to compute the players' expected payoffs. 
Because the population mostly consists of residents, and because residents mutually cooperate with each other, their expected payoff is $\pi_\gtft = b\!-\!c$. 
On the other hand, the defecting mutant only interacts with residents. 
Given the residents' strategy, the mutant receives a benefit~$b$ in the first round, and in every subsequent round with probability $q$. 
As a result, the mutant's expected payoff is $\pi_\alld \!=\! (1\!-\!\delta\!+\!\delta q)b$. 
For perfect payoff memory, the requirement for cooperation to be stochastically stable reduces to the condition $\pi_\gtft > \pi_\alld$, which yields
\begin{equation} \label{Eq:PerfectMemory}
q < 1\!-\!\frac{c}{\delta  b}.
\end{equation}
In particular, we recover the previous observation that $q\!=\!1-c/(\delta b)$ is the maximum generosity that cooperators should have~\citep{molander:jcr:1985,Nowak1992tit,Schmid:NHB:2021}. 
Moreover, because $q$ needs to be non-negative, we also conclude that cooperation can only be stable if $\delta \!>\! c/b$.
Again, this condition for the feasibility of direct reciprocity already exists in the literature~\citep{nowak:Science:2006}.

@@

The logic of the case with limited payoff memory is somewhat different. 
Here we need to compute how likely each player obtains one of the four possible payoffs $\{b\!-\!c, -c, b, 0\}$ in the very last round of a game, before they make social comparisons.
Because residents almost always interact with other residents, their last one-shot payoff is $\pi_\gtft = b\!-\!c$ almost surely. 
For the defecting mutant, there are two possibilities. If the mutant's co-player happens to cooperate in the last round, the mutant receives $\pi_\alld\!=\!b$. This case occurs with probability $1\!-\!\delta\!+\!\delta q$. On the other hand, if the co-player defects in the last round, the mutant receives $\pi_\alld=\!0$. This occurs with the converse probability $\delta(1\!-\!q)$.
Because $b\!-\!c\!<\!b$, residents tend to imitate the mutant in the first case. 
Because $b\!-\!c\!>\!0$, mutants tend to imitate the resident in the second case. 
Cooperation is stochastically stable if the first case is less likely than the second, which yields the condition
\begin{equation} \label{Eq:LimitedMemory}
q < 1\!-\!\frac{1}{2 \delta}.
\end{equation}
Interestingly, this condition no longer depends on the exact payoff values, $b$ and $c$. 
This independence arises because of our assumption of strong selection, in which case only the payoff ordering $b\!>\!c\!>\!0$ matters. Because $q$ is non-negative, condition~\eqref{Eq:LimitedMemory} can be satisfied only if $\delta\!>\!1/2$. That is, players need to interact in more than two rounds, on average.

@@

By comparing the two cases, we find that payoff memory affects whether a conditionally cooperative strategy ($1,1,q$) is viable. 
With perfect memory, the maximum generosity $q$ must satisfy Eq.~\eqref{Eq:PerfectMemory}.
In particular, this generosity can become arbitrarily large, provided the game's benefit-to-cost ratio~$b/c$ and the continuation probability~$\delta$ are sufficiently large. 
In contrast, with limited payoff memory, the maximum generosity is bounded by one half, and it is independent of the benefit-to-cost ratio.\\

@@

\noindent
{\bf Evolutionary dynamics of reciprocity.}
The previous static observations describe whether full cooperation, once established, can be sustained. 
In a next step, we explore in which cases cooperation actually evolves. 
To this end, we turn to simulations.
We have run separate simulations for perfect and limited payoff memory. 
In each case, we consider both a low and a high benefit of cooperation ($b/c\!=\!3$ and $b/c\!=\!10$, respectively).  
For each simulation, we record which strategies ($y,p,q$) the players adopt over time.
\FigBaseResults{} depicts the conditional cooperation probabilities $p$ and $q$ (we omit the opening move \(y\) because we use a discount factor~\(\delta\) close to one, such that first-round behavior is largely irrelevant). 
In all simulations, we find that the players' strategies cluster in two regions of the strategy space. 
The first region corresponds to a neighborhood of \alld{} with $(p,q)\!\approx\!(0,0)$.
The second region corresponds to a thin strip of cooperative strategies with $(p,q)\!\approx\!(1,q)$. 
Within this strip, we observe that most strategies satisfy the constraints on $q$ suggested by the inequalities~\eqref{Eq:PerfectMemory} and~\eqref{Eq:LimitedMemory}. 
That is, with perfect memory, most evolving strategies have $q\!<\!1\!-\!c/b$, whereas with limited payoff memory, most strategies have $q\!<\!1/2$. 
In particular, for limited payoff memory, changes in the benefit parameter have no effect on the qualitative distribution of strategies.

@@

We complement these analytical results with additional simulations; see \FigHigherMemory.
We observe that a minimal increase in the players' payoff memory (compared to the baseline case with a single round recalled) can promote cooperation considerably. 
Specifically, in all four scenarios with extended memory, we see similar cooperation rates that approach the rates observed under perfect memory. 
These results suggest that while it takes {\it some} payoff memory to sustain substantial cooperation rates, the memory requirements are rather modest. 
Already remembering a few interactions, either with the same co-player or across different co-players, may provide players with enough information to adopt reciprocal strategies.\\
